---
title: "Tools for Reproducible Psychometric Experimentation"
format: 
  html:
    toc: true
vignette: >
  %\VignetteIndexEntry{Tools for Reproducible Psychometric Experimentation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| message: false
#| warning: false
library(dropit)
library(psych)
```

We will use the BFI data from the `psych` package to illustrate the `dropit()` function for scale modifications. The BFI is a personality inventory that measures the Big Five traits: agreeableness, conscientiousness, extraversion, neuroticism, and openness. Each trait is assessed with 5 items, resulting in a total of 25 items.

```{r}
# Load BFI data
dta <- psych::bfi

# Get the position of (non) BFI items
xcld <- c("gender", "education", "age") 
xcld_pat <- colnames(dta) %in% xcld

# Store the data in a matrix
bfi <- as.matrix(dta[, !xcld_pat])

# Number of respondents
N <- nrow(bfi)

# Number of items
I <- ncol(bfi)

# Labels
rownames(bfi) <- paste0("p",seq(N))

# Number of dimensions
D <- 5
```

## Assignment

The Q-matrix is a binary matrix that specifies which items measure which construct(s). Here, it helps us define which BFI items belong to which domain. A nice feature of this representation is that it allows easy computation of sum scores for each domain by matrix multiplication. 

```{r}
Q <- matrix(0, nrow = I, ncol = D) 
rownames(Q) <- colnames(bfi)
colnames(Q) <- c("a", "c", "e", "n", "o")

Q[1:5, "a"]   <- 1 # Agreeableness (A1-A5)
Q[6:10, "c"]  <- 1 # Conscientiousness (C1-C5)
Q[11:15, "e"] <- 1 # Extraversion (E1-E5)
Q[16:20, "n"] <- 1 # Neuroticism (N1-N5)
Q[21:25, "o"] <- 1 # Openness (O1-O5)

# Sumscores (original)
bfi %*% Q |> head()
```

## Extraversion Domain

For now, let's focus on the Extraversion items:

```{r}
# Extraversion items
extra <- bfi[,Q[,"e"] == 1]
extra |> head()
```

## Drop It!

What happens when we drop the two worst-performing items from the extraversion scale in a single pass based on their alpha values? The `dropit()` function allows us to explore that question.

```{r}
drop2ga <- dropit(
  data = extra,
  n_drop = 2L,
  direction = "tail",
  criterion = "alpha",
  approach = "oneshot",
  alpha_metric = c("raw_alpha"),
  alpha_args = list(check.keys = TRUE),
  verbose = FALSE 
)

# Manifest correlation between sum scores
cor(rowSums(extra), rowSums(drop2ga$subset), use = "pairwise.complete.obs")
```

## The Dropping Criterion: Alpha vs. Lambda

With the `criterion` argument, the user specifies the metric to score items. The two options are:

* **`criterion = "alpha"` (Coefficient Alpha):** This approach relies on `psych::alpha()`. The algorithm drops the item whose removal *least* reduces (or most increases) the scale's overall Cronbach's alpha.

* **`criterion = "lambda"` (Factor Loading):** This approach relies on `lavaan::cfa()`. It fits a one-factor Confirmatory Factor Analysis model and ranks items based on their absolute standardized factor loadings ($\lambda$). It drops the item with the weakest (or strongest) loadings on the underlying trait.

Let's compare dropping a single item from our extraversion scale using both methods:

```{r}
# Drop weakest item based on Cronbach's alpha
drop_alpha <- dropit(
  data = extra,
  n_drop = 1L,
  criterion = "alpha",
  verbose = FALSE
)

# Drop weakest item based on CFA factor loadings
drop_lambda <- dropit(
  data = extra,
  n_drop = 1L,
  criterion = "lambda",
  verbose = FALSE
)

cat("Alpha:", drop_alpha$names, "\n")
cat("Lambda:", drop_lambda$names, "\n")
```

## Greedy vs. One-Shot

When dropping multiple items (`n_drop > 1`), the approach used to select them becomes more critical. `dropit` offers two approaches:

* **`approach = "oneshot"`:** The algorithm evaluates the full scale, ranks all items simultaneously, and, by default, drops the $n$ items ranked worst under the chosen criterion, in a single pass.
* **`approach = "greedy"`:** The algorithm evaluates the scale and drops only the single worst item. It then *recomputes* alpha or lambda of the remaining items, recalculates the rankings, and drops the next worst item. This repeats $n$ times.

Why does this matter? Well let's compare the two approaches when dropping three items from our extraversion scale using the lambda criterion:

```{r}
# One-shot dropping
drop3_oneshot <- dropit(
  data = extra,
  n_drop = 3L,
  criterion = "lambda",
  approach = "oneshot",
  verbose = FALSE
)

# Greedy dropping
drop3_greedy <- dropit(
  data = extra,
  n_drop = 3L,
  criterion = "lambda",
  approach = "greedy",
  verbose = FALSE
)

cat("One-shot:", paste(drop3_oneshot$names, collapse = ", "), "\n")
cat("Greedy:", paste(drop3_greedy$names, collapse = ", "), "\n")
```

## Adversarial Modifications

While scale abbreviation typically aims to minimize information loss by dropping the weakest items (`direction = "tail"`), `dropit` also allows you to establish worst-case scenarios. Setting `direction = "head"` forces the algorithm to drop the *strongest* items—those with the highest absolute CFA loadings or whose removal most severely reduces Cronbach's alpha.

```{r}
# Best-case abbreviation (lowest)
bestcase <- dropit(
  data = extra,
  n_drop = 2,
  criterion = "lambda",
  direction = "tail",
  verbose = FALSE
)

# Worst-case abbreviation (highest)
worstcase <- dropit(
  data = extra, 
  n_drop = 2, 
  criterion = "lambda", 
  direction = "head", 
  verbose = FALSE
)

cat("Best-case:", paste(bestcase$names, collapse = ", "), "\n")
cat("Worst-case:", paste(worstcase$names, collapse = ", "), "\n")
```

## Retaining Facets

Most domains in modern psychological inventories are multifaceted. If we apply an abbreviation algorithm to the entire dimension at once, it might disproportionately decimate one facet while leaving another completely untouched, destroying the theoretical structure of the scale.

The `partition` argument solves this by applying the dropping algorithm independently across user-defined subscales. Since the built-in `bfi` dataset does not specify official extraversion facets, let's arbitrarily divide our five items into two imaginary sub-components: "S" (E1, E2, E3) and "A" (E4, E5). We can then instruct `dropit()` to drop the single weakest item from each facet simultaneously.

```{r}
# Arbitrarily assign: 5 extraversion items to 2 facets
fcts_extra <- c("S", "S", "S", "A", "A")
# Drop 1 item from each facet
drop_multi <- dropit(
  data = extra,
  partition = fcts_extra,
  n_drop = 1L,
  direction = "tail",
  criterion = "alpha",
  approach = "oneshot",
  alpha_metric = "raw_alpha",
  alpha_args = list(check.keys = TRUE),
  verbose = FALSE
)
# Result is now organized by facet
drop_multi$names
```

## Anchoring Items

Researchers often need to retain specific items to ensure strict comparability against an established baseline. The `anchor` argument allows you to define a set of items that are immune to dropping. Let's assume items `E2` and `E4` are theoretically vital to the extraversion scale. However, in the previous run, the algorithm dropped them. By setting them as anchors, we ensure they remain protected.

```{r}
drop2ga_anc <- dropit(
  data = extra,
  anchor = c("E2", "E4"),
  n_drop = 2L,
  direction = "tail",
  criterion = "alpha",
  approach = "oneshot",
  alpha_metric = "raw_alpha",
  alpha_args = list(check.keys = TRUE),
  verbose = FALSE
)$names

# Compare dropped items against unanchored baseline
cat("Unanchored:", paste0(drop2ga$names, collapse = ", "), "\n")
cat("Anchored:", paste0(drop2ga_anc, collapse = ", "), "\n")
```

## Missing Data

Missing data are a common problem in most empirical datasets, and how they are handled can substantially affect the results of any procedure or analysis. `dropit()` does not have its own approach to missing data handling; instead, it hands these values straight to the ranking engine via two arguments: `alpha_args` for `psych::alpha()` and `cfa_args` for `lavaan::cfa()`. The choice should stay visible to the user rather than hidden behind a wrapper option—and, because the two engines are configured through their own native arguments, it is plain to any reader that the alpha path and the lambda path handle missingness in different ways.

### Alpha Dropping with Missing Data

When using `criterion = "alpha"`, the internal rankings are calculated via `psych::alpha()`. By default, this function handles missing values by using pairwise deletion (`use = "pairwise"`) to construct the underlying correlation matrix. However, if you want to enforce complete cases (listwise deletion), you can pass the `use` argument directly into the `alpha_args` list:

```{r}
drop2oa_co <- dropit(
  data = extra,
  n_drop = 2,
  criterion = "alpha",
  approach = "oneshot",
  verbose = FALSE,
  alpha_args = list(
    check.keys = TRUE, 
    use = "complete.obs"  # change strategy
  )
)
```

### Lambda Dropping with Missing Data

With `criterion = "lambda"`, the internal CFA models are fitted using `lavaan::cfa()`, which defaults to **listwise deletion**. If your dataset contains scattered missing values, you might want to use Full Information Maximum Likelihood (FIML) instead. Because `dropit` acts as a seamless wrapper, you can pass `missing = "fiml"` directly into `cfa_args`:

```{r}
drop2_fiml <- dropit(
  data = extra,
  n_drop = 2,
  criterion = "lambda",
  approach = "greedy",
  verbose = FALSE,
  cfa_args = list(
    std.lv = TRUE,
    missing = "fiml"  # full information maximum likelihood
  )
)
```

### Greedy Dropping and Sample Stability

There is one combination worth pausing on: greedy dropping on data that still contains missing values. Recall that the greedy approach refits the model after every removal. Each of those refits runs on a smaller set of items, so the engine re-derives its working sample each round from whichever rows are usable for the items that remain. With scattered missingness, that sample can quietly shift from one round to the next. Because this happens inside the loop, it leaves no trace in the returned object.

`dropit()` therefore raises a warning whenever `approach = "greedy"` meets a dataset with missing values, so the drift is stated rather than silent. As with every warning `dropit()` collects, it is retained on the returned object and reachable through `$log`:

```{r}
# Introduce some scattered missingness
extra_na <- extra
extra_na[sample(nrow(extra_na), 20), "E2"] <- NA

drop2_drift <- dropit(
  data = extra_na,
  n_drop = 2,
  criterion = "lambda",
  approach = "greedy",
  verbose = FALSE
)

# The warning is kept in the log
drop2_drift$log$warnings
```

If a fixed sample across rounds matters—and for a comparison against a baseline it usually does—treat the missingness *before* the call. In the simplest case, you can just restrict the dataset to complete cases.

```{r}
drop2_stable <- dropit(
  data = extra_na[complete.cases(extra_na), ],
  n_drop = 2,
  criterion = "lambda",
  approach = "greedy",
  verbose = FALSE
)

cat("Dropped:", paste(drop2_stable$names, collapse = ", "), "\n")
```

## Simulation Baselines

Because `dropit` is designed for experimental modifications, users might want to compare abbreviated scales against the full, unmodified scale. To facilitate seamless programmatic loops, passing `n_drop = 0` acts as a "no-operation." It bypasses the dropping algorithms and safely returns the original dataset wrapped in the standard `dropit` object structure. This prevents automated pipelines from crashing at the zero-drop baseline.

```{r}
# Dropping 0-3 items
res <- lapply(0:3, function(k) {
  dropit(
    data = extra, 
    n_drop = k, 
    criterion = "lambda", 
    verbose = FALSE
  )
})

# Extract retained datasets
base <- lapply(res, `[[`, "subset")

# View dims to confirm decay
sapply(base, dim)
```

Inside a loop like this, `dropit()` re-runs its advisory guards—the boundary-tie, small-scale, and greedy-with-missing-data warnings—on every single iteration, quietly re-collecting the same messages into each `$log`. Once you have validated your design and know those advisories no longer tell you anything new, `checks = FALSE` switches them off for the whole run. Note that `verbose = FALSE` will *not* do this: it only hides the printed summary, while the guards still fire and still land in the log. The switch changes nothing about which items are dropped—only whether the guards run.

```{r}
res <- lapply(0:3, function(k) {
  dropit(
    data = extra,
    n_drop = k,
    criterion = "lambda",
    checks = FALSE,  # skip the advisory guards across the simulation
    verbose = FALSE
  )
})

sapply(lapply(res, `[[`, "subset"), dim)
```

## Reproducibility

If your abbreviation strategy relies on stochastic `lavaan` arguments (e.g., bootstrap standard errors or random starting values), you can pass a `seed` argument directly to `dropit()`. Crucially, `dropit` takes a snapshot of your global random state (`.Random.seed`), applies your exact seed for the internal mechanics, and then safely restores the original state upon exit. This guarantees reproducibility for the abbreviation baselines without permanently side-effecting your R session's broader simulation environment.

```{r}
# Run CFA with bootstrapping twice; same seed
res1 <- dropit(
  data = extra, 
  n_drop = 1L, 
  criterion = "lambda", 
  seed = 42, 
  # Note: bootstrap iterations kept intentionally low for vignette compilation speed
  cfa_args = list(se = "bootstrap", bootstrap = 5), 
  verbose = FALSE
)

res2 <- dropit(
  data = extra, 
  n_drop = 1L, 
  criterion = "lambda", 
  seed = 42, 
  cfa_args = list(se = "bootstrap", bootstrap = 5),
  verbose = FALSE
)

# Verify subsets are identical
identical(res1$subset, res2$subset)
```
