---
title: "Tools for Reproducible Psychometric Experimentation"
format: 
  html:
    toc: true
vignette: >
  %\VignetteIndexEntry{Tools for Reproducible Psychometric Experimentation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| message: false
#| warning: false
library(dropit)
library(psych)
```

We will use the BFI data from the `psych` package to illustrate the `dropit()` function for item dropping based on Cronbach's alpha.

```{r}
# Load BFI data
dta <- psych::bfi

# Get the position of (non) BFI items
xcld <- c("gender", "education", "age") 
xcld_pat <- colnames(dta) %in% xcld

# Store the data in a matrix
bfi <- as.matrix(dta[, !xcld_pat])

# Number of respondents
N <- nrow(bfi)

# Number of items
I <- ncol(bfi)

# Labels
rownames(bfi) <- paste0("p",seq(N))

# Number of dimensions
D <- 5
```

## Assignment

The Q-matrix is a binary matrix that specifies the mapping of an item to its respective latent construct(s). Here, it helps us define which BFI items belong to which domain. A nice feature of this representation is that it allows easy computation of sum scores for each domain by matrix multiplication. 

```{r}
Q <- matrix(0, nrow = I, ncol = D) 
rownames(Q) <- colnames(bfi)
colnames(Q) <- c("a", "c", "e", "n", "o")

Q[1:5, "a"]   <- 1 # Agreeableness (A1-A5)
Q[6:10, "c"]  <- 1 # Conscientiousness (C1-C5)
Q[11:15, "e"] <- 1 # Extraversion (E1-E5)
Q[16:20, "n"] <- 1 # Neuroticism (N1-N5)
Q[21:25, "o"] <- 1 # Openness (O1-O5)

# Sumscores (original)
bfi %*% Q |> head()
```

## Extraversion Domain

For now, let's focus on the Extraversion items:

```{r}
# Extraversion items
extra <- bfi[,Q[,"e"] == 1]
extra |> head()
```

## Drop It!

What happens, when we greedily drop, say the 2 worst-performing items from the extraversion scale based on their alpha values? The `dropit()` function allows us to explore that question.

```{r}
drop2ga <- dropit(
  data = extra,
  n_drop = 2L,
  direction = "tail",
  criterion = "alpha",
  approach = "greedy",
  alpha_metric = c("raw_alpha"),
  alpha_args = list(check.keys = TRUE),
  verbose = FALSE
)

# Manifest correlation between sum scores
cor(rowSums(extra), rowSums(drop2ga$subset), use = "pairwise.complete.obs")
```

## Retaining Facets

Most domains in modern psychological inventories are multifaceted. If we apply an abbreviation algorithm to the entire dimension at once, it might disproportionately decimate one facet while leaving another completely untouched, destroying the theoretical structure of the scale.

The partition argument solves this by applying the dropping algorithm independently across user-defined subscales. Since the built-in bfi dataset does not specify official Extraversion facets, let's arbitrarily divide our five items into two imaginary sub-components: "S" (E1, E2, E3) and "A" (E4, E5). We can then instruct `dropit()` to drop the single weakest item from each facet simultaneously.

```{r}
# Arbitrarily assign the 5 Extraversion items to two facets
extra_facets <- c("S", "S", "S", "A", "A")
# Drop 1 item from each facet
drop_multi <- dropit(
  data = extra,
  partition = extra_facets,
  n_drop = 1L,
  direction = "tail",
  criterion = "alpha",
  approach = "oneshot",
  alpha_metric = "raw_alpha",
  alpha_args = list(check.keys = TRUE),
  verbose = FALSE
)
# The result is now organized by facet
drop_multi$names
```

## Anchoring Items

Researchers often need to retain specific items to ensure strict comparability against an established baseline. The `anchor` argument allows you to define a set of items that are immune to dropping. Let's assume items `E2` and `E4` are theoretically vital to the Extraversion scale. However, in the previous run, the algorithm dropped them. By setting them as anchors, we ensure they remain protected.

```{r}
drop2ga_anc <- dropit(
  data = extra,
  anchor = c("E2", "E4"),
  n_drop = 2L,
  direction = "tail",
  criterion = "alpha",
  approach = "greedy",
  alpha_metric = "raw_alpha",
  alpha_args = list(check.keys = TRUE),
  verbose = FALSE
)$names

# Compare dropped items against unanchored baseline
cat("Unanchored:", paste0(drop2ga$names, collapse = ", "), "\n")
cat("Anchored:", paste0(drop2ga_anc, collapse = ", "), "\n")
```

## Missing Data

Missing data is a common issue in psychometric datasets, and how it is handled can significantly impact the results of item dropping procedures. The `dropit()` function provides flexibility in managing missing data through the `alpha_args` and `cfa_args` parameters, which allow you to specify the method for handling missing values when calculating Cronbach's alpha or fitting CFA models, respectively.

### Alpha Dropping with Missing Data

When using `criterion = "alpha"`, the internal rankings are calculated via `psych::alpha()`. By default, this function handles missing values by using pairwise deletion (`use = "pairwise"`) to construct the underlying correlation matrix. For most survey data, this default is methodologically sound and requires no extra configuration.

However, if you want to strictly enforce complete cases (listwise deletion) to ensure the alpha is computed on the exact same respondents for every item, you can pass the `use` argument directly into the `alpha_args` list:

```{r}
drop2oa_co <- dropit(
  data = extra,
  n_drop = 2,
  criterion = "alpha",
  approach = "oneshot",
  verbose = FALSE,
  alpha_args = list(
    check.keys = TRUE, 
    use = "complete.obs"  # change strategy
  )
)
```

### Lambda Dropping with Missing Data

Missing data handling requires careful attention when using `criterion = "lambda"`. The internal CFA models are fitted using `lavaan::cfa()`, which defaults to **listwise deletion**. If your dataset contains scattered missing values, this default will drop any participant who skipped even a single item, severely reducing your sample size before the factor loadings are estimated.

To prevent this and retain your full dataset, you should leverage Full Information Maximum Likelihood (FIML). Because `dropit` acts as a seamless wrapper, you can pass `missing = "fiml"` directly into `cfa_args`:

```{r}
drop2_fiml<- dropit(
  data = extra,
  n_drop = 2,
  criterion = "lambda",
  approach = "greedy",
  verbose = FALSE,
  cfa_args = list(
    std.lv = TRUE, 
    missing = "fiml"  # full information maximum likelihood
  )
)
```
