---
title: "Tools for Reproducible Psychometric Experimentation"
format: 
  html:
    toc: true
vignette: >
  %\VignetteIndexEntry{Tools for Reproducible Psychometric Experimentation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
library(dropit)
library(psych)
```

We will use the BFI data from the `psych` package to illustrate the `dropit()` function for item dropping based on Cronbach's alpha.

```{r}
# Load BFI data
dta <- psych::bfi

# Get the position of (non) BFI items
xcld <- c("gender", "education", "age") 
xcld_pat <- colnames(dta) %in% xcld

# Store the data in a matrix
bfi <- as.matrix(dta[, !xcld_pat])

# Numer of respondents
N <- nrow(bfi)

# Number of items
I <- ncol(bfi)

# Labels
rownames(bfi) <- paste0("p",seq(N))

# Number of dimensions
D <- 5
```

# Assignment

The Q-matrix is a binary matrix that specifies the mapping of an item to its respective latent construct(s). Here, it helps us define which BFI items belong to which domain. A nice feature of this representation is that it allows easy computation of sum scores for each domain by matrix multiplication. 

```{r}
Q <- matrix(0, nrow = I, ncol = D) 
rownames(Q) <- colnames(bfi)
colnames(Q) <- c("a", "c", "e", "n", "o")

Q[1:5, "a"] <- 1 # Extraversion
Q[6:10, "c"] <- 1 # Agreeableness
Q[11:15, "e"] <- 1 # Conscientiousness
Q[16:20, "n"] <- 1 # Neuroticism
Q[21:25, "o"] <- 1 # Openness

# Sumscores (original)
bfi %*% Q |> head()
```

# Extraversion Domain

For now, let's focus on the Extraversion items:

```{r}
# Extraversion items
extra <- bfi[,Q[,"e"] == 1]
extra |> head()
```

# Drop It!

What happens, when we greadily drop, say the 2 worst-performing items from the extraversion scale based on their alpha values? The `dropit()` function allows us to explore that question.

```{r}
drop2ga <- dropit(
  extra,
  n_drop = 2L,
  direction = "tail",
  criterion = "alpha",
  approach = "greedy",
  output_type = "both",
  alpha_metric = c("raw_alpha"),
  alpha_args = list(check.keys = TRUE),
  verbose = FALSE
)
drop2ga$names

# Manifest correlation between sum scores
cor(rowSums(extra), rowSums(drop2ga$subset), use = "pairwise.complete.obs")
```

# Anchoring Items

Researchers often need to retain specific items to ensure strict comparability against an established baseline. The `anchor` argument allows you to define a set of items that are immune to dropping. Let's assume items `E2` and `E4` are theoretically vital to the Extraversion scale. However, in the previous run, the algorithm dropped them. By setting them as anchors, we ensure they remain protected.

```{r}
drop2_anc <- dropit(
  extra,
  anchor = c("E2", "E4"),
  n_drop = 2L,
  direction = "tail",
  criterion = "alpha",
  approach = "greedy",
  output_type = "names",
  alpha_metric = "raw_alpha",
  alpha_args = list(check.keys = TRUE),
  verbose = FALSE
)

# Compare dropped items against unanchored baseline
cat("Unanchored:", paste0(drop2ga$names, collapse = ", "), "\n")
cat("Anchored:", paste0(drop2_anc, collapse = ", "), "\n")
```